---
title: "Multiway Classification with Real Data"
author: "Matthew Asisgress"
date: "2025-11-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We apply R package **cpfa** (Asisgress, 2025) to four real datasets. First, we 
apply the package to MNIST (LeCun, Cortes, and Burges, 1998; LeCun et al., 
2002)---showing how the package can be used to distinguish between images of the
digits 2 and 3. Second, we apply the package to Fashion MNIST (Xiao, Rasul, and 
Vollgraf, 2017)---distinguishing among images of tops, trousers, and sandals. 
Third, we apply the package to VesselMNIST3D (Yang et al., 2020) from the 
MedMNIST database (Yang et al., 2023; Yang, Shi, Ni, 2021). VesselMNIST3D 
contains three-dimensional representations of blood vessels, and we distinguish 
healthy blood vessels from aneurysm blood vessels. Finally, we apply the package 
to OrganMNIST3D (Bilic et al., 2023; Xu et al., 2019) from the MedMNIST database 
(Yang et al., 2023; Yang, Shi, Ni, 2021). OrganMNIST3D contains 
three-dimensional representations of abdominal organs, and we distinguish among 
four organs.

# A.) MNIST

MNIST consists of 70,000 grayscale images of handwritten digits from 0 to 9. We 
use R package **dslabs** (Irizarry and Gill, 2025) to download the dataset. We 
subset to only 2000 images from the original training set. The images include 
980 images of the digit 2 and 1020 images of the digit 3.

## Download

```{r, message = FALSE, warning = FALSE}
# load library
library(dslabs)

# download MNIST data and subset training set to digits of 2 or 3
mnist <- read_mnist()
inde <- which(mnist$train$labels %in% c(2, 3))
images <- mnist$train$images[inde, ]
labels <- mnist$train$labels[inde]

# restructure data into three-way array and prepare labels
X0 <- array(images, c(nrow(images), 28, 28))
y0 <- as.factor(as.numeric(as.factor(labels)) - 1)

# subset to first 2000 images
ind <- 1:2e3
X <- X0[ind, , ]
y <- y0[ind]
```

We plot an example of digit 2 and digit 3.

```{r}
# plot example of 2 and 3
par(mfrow = c(1, 2))
pdigit <- function(imat) {
  m <- t(apply(imat, 2, rev))
  image(m, col = gray(seq(0, 1, 0.05)), xaxt = "n", yaxt = "n")
}
for (i in 0:1) {pdigit(t(X[which(y == i)[1], , ]))}
```

```{r echo = FALSE}
par(mfrow = c(1, 1))
```

## Analysis

We load R package **cpfa** and initialize the tensor model. First, the data 
array is a regular three-way array; so we set `model <- "parafac"` to use a 
Parafac model (Harshman, 1970). Second, we initialize the number of components 
to fit for the model by setting `nfac <- c(2, 3)` in order to fit both a 
two-component Parafac model and a three-component Parafac model. We set 
`nstart <- 10` to allow for 10 random starts in the Parafac alternating least 
squares algorithm fit by R package **multiway** (Helwig, 2025), upon which R 
package **cpfa** depends. Third, we specify the constraint desired for each 
array mode using `const`. Fourth, we use `cmode <- 1` to specify that the 
classification mode is the first mode of the input array (i.e., the mode 
connected to the class labels). Note that numerous constraint options are 
available; after loading **cpfa**, type `const()` in the R console to access a 
constraint options list provided by R package **CMLS** (Helwig, 2025). 

Next, we initialize classification methods. First, we use 
`method = c("PLR", "RF")` to employ penalized logistic regression (PLR) and 
random forest (RF) classifiers for this problem. See `help(cpfa)` for 
information on additional classification methods. Second, we specify that the 
problem is a binary classification problem by setting `family <- "binomial"`. 
Third, we use 10-fold cross-validation (CV) in our inner training, setting 
`nfolds <- 10`; and fourth, we set `nrep <- 5` to perform five outer train-test 
splits of our data. Fifth, we set `ratio <- 0.9` to specify that each outer 
training set contains a proportion of 0.9 of the full input data while the outer 
testing set contains a proportion of 0.1. Finally, we specify ranges for tuning 
parameters alpha (PLR), the number of trees (RF), and node size (RF), wrapping 
them into the list called `parameters`.

```{r, message = FALSE, warning = FALSE}
# load library
library(cpfa)

# set seed
set.seed(500)

# initialize model 
model <- "parafac"
nfac <- c(2, 3)
nstart <- 10
const <- c("uncons", "uncons", "uncons")
cmode <- 1

# initialize classification
method <- c("PLR", "RF")
family <- "binomial"
nfolds <- 10
nrep <- 5
ratio <- 0.9

# initialize tuning parameters
alpha <- seq(0, 1, length = 8)
ntree <- c(400, 600, 800, 1000)
nodesize <- c(4, 8, 16, 32)
parameters <- list(alpha = alpha, ntree = ntree, nodesize = nodesize)

# implement train-test splits with inner k-fold CV to optimize classification
outputR <- cpfa(x = X, y = y, model = model, nfac = nfac, nstart = nstart,
                const = const, cmode = cmode, method = method, family = family, 
                nfolds = nfolds, nrep = nrep, ratio = ratio, 
                parameters = parameters, type.out = "descriptives", 
                seeds = NULL, plot.out = TRUE, parallel = FALSE, 
                verbose = FALSE)

```

## Results

We examine classification performance metrics error (`err`) and accuracy (`acc`) 
for each model and for each classifier, looking at median metrics across outer 
train-test splits. We also examine, averaged across train-test splits, optimal 
tuning parameters chosen for each classifier.

```{r}
# examine classification performance measures - median across train-test splits
outputR$descriptive$median[, 1:2]

# examine optimal tuning parameters - mean across train-test splits
outputR$mean.opt.tune
```

We see that accuracy values are close to 0.9 for both two-component and 
three-component models. The two-component Parafac model with the RF classifier 
performed best. In addition, we can see average optimal tuning parameters for 
each model. Note that PLR optimized lambda internally.

Next, we use function `plotcpfa`. This function performs two tasks: (1) it 
fits the optimal model among those fit using function `cpfa`; and (2) it plots 
estimated component weights from that model for non-classification modes. 
Looking at such plots, we can search for relationships between model components 
and levels of mode A or B. In this case, mode A corresponds to the horizontal 
axis of MNIST images while mode B corresponds to the vertical axis.

```{r}
# set seed
set.seed(500)

# plot heatmaps of component weights for optimal model
results <- plotcpfa(outputR, nstart = 10, ctol = 1e-6, verbose = FALSE)
```

For the A mode, component weights are stronger between levels 5 and 24 for the 
first component. Looking at the B mode, weights are stronger between levels 4 
and 25. Whether viewed from mode A or B the first component appears to be a 
general component identifying the presence of non-zero values, distinguishing 
empty outer parts of images from non-zero inner parts. The digits 2 and 3 have 
systematic differences in the presence of non-zero values in these areas.

Examining the second component for the A mode, weights are stronger around level 
14 where there appears to be a greater presence of zero values in digit 3, 
compared to digit 2. A similar pattern exists in mode B around levels 18 and 19 
with more non-zero values for 2, compared to 3. Taken together between the two 
modes, this component might be identifying the closed circle present in the 
digit 2, which differs from the open or half circle seen in digit 3.

# B.) Fashion MNIST

Fashion MNIST consists of 70,000 grayscale images of fashion objects within 10 
different categories, indexed by class labels of 0 through 9 (Xiao, Rasul, and 
Vollgraf, 2017). We use R package **keras3** (Kalinowski, Allaire, and Chollet, 
2025) to download the dataset. We subset to 1000 images from the training set. 
We only use images within the categories of top (with a label of 0), trouser 
(1), or sandal (5). Images include 308 tops, 357 trousers, and 335 sandals. 

We remove a different number of horizontal levels from each image, randomly 
selecting a number of levels between three and six (inclusive) to remove for 
each image. The resulting array is ragged: mode A (the horizontal mode) contains 
a different number of rows for each image while mode B (the vertical mode) 
contains exactly 28 levels for all images. This removal is meant to simulate 
data corruption or sensor malfunction that could lead to incomplete images. To 
continue forward, we use a Parafac2 model (Harshman, 1972), which can be fit to 
a ragged array (i.e., one mode has a different number of levels, conditional on 
another mode). Parafac2 maintains properties similar to the Parafac model, 
including the intrinsic axis property (for details, see Harshman and Lundy, 
1994, 1996).

## Download

```{r, message = FALSE, warning = FALSE}
# load library
library(keras3)

# download Fashion MNIST and subset training set to categories of 0, 1, 5
fmnist <- dataset_fashion_mnist()
inde <- which(fmnist$train$y %in% c(0, 1, 5))
X0 <- fmnist$train$x[inde, ,]
labels <- fmnist$train$y[inde]

# prepare labels by converting to class factor with levels of 0, 1, and 2
y0 <- as.factor(as.numeric(as.factor(labels)) - 1)

# subset to first 1000 images
ind <- 1:1e3
nimage <- length(ind)
Xp <- X0[ind, , ]
y <- y0[ind]

# permute array to make third mode classification mode 
X <- aperm(Xp, c(2, 3, 1))

# set seed
set.seed(500)

# remove rows from each image and restructure array into list of images
nremove <- sample(3:6, nimage, replace = TRUE)
Xrag <- vector(mode = "list", length = nimage)
for (i in 1:nimage) {
   ranremove <- c(sample(1:28, nremove[i]))
   image0 <- X[-ranremove, , i]
   Xrag[[i]] <- image0
}
```

We plot an example of a top, trouser, and sandal.

```{r}
# plot example of top, trouser, and sandal
par(mfrow = c(1, 3))
pdigit <- function(imat) {
  m <- t(apply(imat, 2, rev))
  image(m, col = gray(seq(0, 1, 0.05)), xaxt = "n", yaxt = "n")
}
for (i in 0:2) {pdigit(Xrag[[which(y == i)[1]]])}
```

```{r echo = FALSE}
par(mfrow = c(1, 1))
```

## Analysis

We load R package **cpfa** and initialize the tensor model. First, the data 
array is a ragged, irregular three-way array; so we set `model <- "parafac2"` to 
use a Parafac2 model. Second, we initialize the number of components to fit by 
setting `nfac <- c(2, 3)`. Third, we set `nstart <- 10` for 10 random starts. 
Fourth, we specify the constraint for each mode using `const`. For Parafac2, we 
set the third mode to have non-negative weights.

Next, we initialize classification methods. First, we use 
`method = c("SVM", "GBM")` to use support vector machine (SVM) and gradient 
boosting machine (GBM). Second, we specify that the problem is a multiclass 
classification problem by setting `family <- "multinomial"`. Third, we use 
10-fold CV in our inner training, setting `nfolds <- 10`; and fourth, we set 
`nrep <- 5`. Fifth, we set `ratio <- 0.9`. Finally, we specify ranges for tuning 
parameters gamma (SVM), cost (SVM), eta (GBM), maximum depth (GBM), subsample 
(GBM), and number of rounds (GBM), wrapping them into `parameters`.

```{r, message = FALSE, warning = FALSE}
# load library
library(cpfa)

# set seed
set.seed(500)

# initialize model 
model <- "parafac2"
nfac <- c(2, 3)
nstart <- 10
const <- c("uncons", "uncons", "nonneg")

# initialize classification
method <- c("SVM", "GBM")
family <- "multinomial"
nfolds <- 10
nrep <- 5
ratio <- 0.9

# initialize tuning parameters
gamma <- c(0, 0.1, 1, 10); cost <- c(0.1, 1, 10, 100)
eta <- c(0.3, 0.7); max.depth <- c(1, 2)
subsample <- c(0.75, 0.9); nrounds <- c(100, 400)
parameters <- list(gamma = gamma, cost = cost, eta = eta, max.depth = max.depth, 
                   subsample = subsample, nrounds = nrounds)

# implement train-test splits with inner k-fold CV to optimize classification
outputR2 <- cpfa(x = Xrag, y = y, model = model, nfac = nfac, nstart = nstart,
                 const = const, method = method, family = family, 
                 nfolds = nfolds, nrep = nrep, ratio = ratio, 
                 parameters = parameters, type.out = "descriptives", 
                 seeds = NULL, plot.out = TRUE, parallel = FALSE, 
                 verbose = FALSE)
```

## Results

We examine classification performance metrics for each model and for each 
classifier, this time averaged across train-test splits. We also examine, 
averaged across train-test splits, the optimal tuning parameters chosen for each 
classifier.

```{r}
# examine classification performance measures - mean across train-test splits
outputR2$descriptive$mean[, 1:2]

# examine optimal tuning parameters - mean across train-test splits
outputR2$mean.opt.tune
```

Accuracy values are between 0.9 and 0.928. The three-component Parafac2 model 
with the GBM classifier performed best. In addition, average optimal tuning 
parameters are available for each model. 

Next, we use function `plotcpfa`. Because the number of levels differs among 
images for mode A, we cannot inspect mode A weights easily. Function `plotcpfa` 
automatically excludes mode A weights from plots based on the Parafac2 model. 
Instead, we inspect only the component weights for mode B to search for 
relationships between the levels of B and the Parafac2 model components.

```{r}
# set seed
set.seed(500)

# plot heatmaps of component weights for optimal model
results2 <- plotcpfa(outputR2, nstart = 10, ctol = 1e-6, verbose = FALSE)
```

For the B mode (i.e., vertical axis), for the first component, weights are 
stronger between levels 11 to 14 and between 16 and 19. This first component 
might be capturing non-zero changes displayed at the center of the top images. 
The first component might help distinguish tops from trousers and sandals.

For the second component, weights are stronger between levels 6 to 23. This 
component could be a general component distinguishing non-zero values in the 
middle columns of images from zero values at the far left and far right columns. 
While the sandal has some non-zero values in the far left and far right columns, 
the top and trouser generally do not; so this second component might help 
distinguish the top and the trouser from the sandal.

For the third component, weights are stronger between levels 6 and 9 and between 
22 and 25. This third component might be capturing zero values to the left and 
right of images, contrasting them with non-zero values in the middle. 
Interestingly, we see two negative bands between levels 11 and 14 and between 16 
and 19; however, level 15 is positive. Level 15 might correspond to the empty 
space between the trouser legs while the negative bands might represent non-zero 
values in the legs. If so, this component might help distinguish trousers from 
tops or sandals.

These interpretations are limited. For all components, more work would need to 
be done to explore the estimated A mode weights (or C mode weights) to 
understand the components from other perspectives. Such work could clarify the 
components further and change the current interpretations.

# C.) VesselMNIST3D from MedMNIST

VesselMNIST3D consists of 1,908 three-dimensional models of brain blood vessels 
based on reconstructed magnetic resonance imaging data. Vessels exist in two 
categories: healthy vessels (indexed by 0) or aneurysm vessels (1). We download 
the dataset manually from <https://medmnist.com/>, which gives us a .npz file. 
We use the Python script `convert_npz_to_h5.py` to convert VesselMNIST3D from 
a file format of .npz to .h5. This script can be found on the project's
repository. Next, we use R package **rhdf5** (Fischer, Smith, and Pau, 2025) to 
read the data (as a .h5) into R. We subset to only 300 objects from the training 
set, including 150 healthy vessels and 150 aneurysm vessels.

## Download

```{r, message = FALSE, warning = FALSE}
# load library
library(rhdf5)

# import VesselMNIST3D data
h5file <- "vesselmnist3d.h5"
imag4d <- h5read(h5file, name = "train_images")
labels <- h5read(h5file, name = "train_labels")

# subset to first 300 objects
ind <- 1:300
Xp <- imag4d[ , , , ind]
y <- as.factor(as.numeric(labels))[ind]

# convert array to double
storage.mode(Xp) <- "double"

# remove full dataset
rm(imag4d)
```

We use R package **plotly** (Sievert, 2020) to generate a point cloud 
representation of a healthy blood vessel.

```{r, message = FALSE, warning = FALSE}
# load library
library(plotly)

# prepare data from vessel 205 for plotting
voxel_df <- as.data.frame(which(Xp[, , , 205] > 0, arr.ind = TRUE))
colnames(voxel_df) <- c("x", "y", "z")

# generate 3D plot
scene1 <- list(title = "", showticklabels = FALSE, showgrid = TRUE, 
               showline = FALSE, zeroline = FALSE)
plot_ly(voxel_df, x = ~x, y = ~y, z = ~z, type = "scatter3d", mode = "markers",
        marker = list(size = 3.5, color = "blue", opacity = 0.3)) %>%
          layout(title = "3D Blood Vessel Example - Point Cloud Representation", 
                 scene = list(xaxis = scene1, yaxis = scene1, zaxis = scene1))
```

## Analysis

We load **cpfa** and initialize the tensor model. First, the data array is a 
regular four-way array; so we set `model <- "parafac"`. Second, we initialize 
the number of components to fit by setting `nfac <- c(2, 3)`. Third, we set 
`nstart <- 10`. Fourth, we specify the constraint for each mode using `const`. 
For Parafac, we set the fourth mode to have non-negative weights.

Next, we initialize classification methods. First, we use 
`method = c("NN", "RDA")` to use feed-forward neural networks (NN) and 
regularized discriminant analysis (RDA). Second, we specify that the problem is 
a binary classification problem, setting `family <- "binomial"`. Third, 
we use 10-fold CV in our inner training, setting `nfolds <- 10`; and fourth, we 
set `nrep <- 5` to perform five outer train-test splits. Fifth, we set 
`ratio <- 0.9`. Finally, we specify ranges for tuning parameters alpha (RDA), 
delta (RDA), size (NN), and decay (NN), wrapping them into `parameters`.

```{r, message = FALSE, warning = FALSE}
# load library
library(cpfa)

# set seed
set.seed(500)

# initialize model 
model <- "parafac"
nfac <- c(2, 3)
nstart <- 10
const <- c("uncons", "uncons", "uncons", "nonneg")

# initialize classification
method <- c("NN", "RDA")
family <- "binomial"
nfolds <- 10
nrep <- 5
ratio <- 0.9

# initialize tuning parameters
rda.alpha <- seq(0, 0.999, length = 4); delta <- c(0, 0.1, 1, 2)
size <- c(1, 2, 4, 8); decay <- c(0.001, 0.01, 0.1, 1)
parameters <- list(rda.alpha = rda.alpha, delta = delta, size = size, 
                   decay = decay)

# implement train-test splits with inner k-fold CV to optimize classification
outputR3 <- cpfa(x = Xp, y = y, model = model, nfac = nfac, nstart = nstart,
                const = const, method = method, family = family, 
                nfolds = nfolds, nrep = nrep, ratio = ratio, 
                parameters = parameters, type.out = "descriptives", 
                seeds = NULL, plot.out = TRUE, parallel = FALSE, 
                verbose = FALSE)
```

## Results

We examine classification performance metrics for each model and classifier, 
averaged across train-test splits. We also examine, averaged across train-test 
splits, optimal tuning parameters for each model.

```{r}
# examine classification performance measures - mean across train-test splits
outputR3$descriptive$mean[, 1:2]

# examine optimal tuning parameters - mean across train-test splits
outputR3$mean.opt.tune
```

Accuracy values are between 0.65 and 0.71. The three-component Parafac model 
with the RDA classifier performed best. In addition, average optimal tuning 
parameters are available for each classifier. 

Next, we use `plotcpfa`. Looking at plots, we can search for relationships 
between the levels of mode A, B, or C and model components. In this case, the 
three modes correspond to the three dimensions of the vessels.

```{r}
# set seed
set.seed(500)

# plot heatmaps of component weights for optimal model
results3 <- plotcpfa(outputR3, nstart = 10, ctol = 1e-6, verbose = FALSE)
```

We omit interpretations of the three components, given the difficulty 
of visualizing each blood vessel object. For more information about this dataset
and classification work using it, consider exploring Yang, Xia, Kin, and 
Igarashi (2020); Yang et al. (2023); and Yang, Shi, and Ni (2021).

# D.) OrganMNIST3D from MedMNIST

OrganMNIST3D consists of 1,742 three-dimensional models of computed tomography 
scans of abdominal organs from the Liver Tumor Segmentation Benchmark (LiTS; 
Bilic et al., 2023). Yang et al. (2023) used bounding-box annotations of 11 body 
organs from another study (Xuanang  et al., 2019) to obtain organ labels. We 
use four categories: bladder (indexed by 0), left femur (1), pancreas (9), and 
spleen (10). As with VesselMNIST3D, we download OrganMNIST3D manually from <https://medmnist.com/>, which gives us a .npz file. After changing the name
from `vesselmnist3d` to `organmnist3d` at lines four and five, we use the 
Python script `convert_npz_to_h5.py` to convert OrganMNIST3D from a file format 
of .npz to .h5. Next, we use R package **rhdf5** (Fischer, Smith, and Pau, 2025) 
to read the data (as a .h5) into R. We subset to only 460 objects from the 
training set---including 115 bladder, 115 left femur, 115 pancreas, and 115 
spleen.

## Download

```{r, message = FALSE, warning = FALSE}
# load library
library(rhdf5)

# import OrganMNIST3D data
h5file <- "organmnist3d.h5"
imag4d <- h5read(h5file, name = "train_images")
labels <- h5read(h5file, name = "train_labels")

# convert labels to factor
y0 <- as.factor(as.numeric(labels))

# identify labels of 0, 1, 9, and 10
ind <- which(y0 %in% c(0, 1, 9, 10))

# subset to bladder, left femur, pancreas, and spleen
Xp <- imag4d[ , , , ind]
y1 <- y0[ind]
y <- droplevels(y1)

# reset class labels
levels(y) <- c(0, 1, 2, 3)

# convert array to double
storage.mode(Xp) <- "double"

# remove full dataset
rm(imag4d)
```

We use R package **plotly** (Sievert, 2020) to generate a colored point cloud 
representation of a spleen.

```{r, message = FALSE, warning = FALSE}
# load library
library(plotly)

# obtain coordinates of voxels above threshold
voxcord <- which(Xp[, , , 459] > 0, arr.ind = TRUE)

# combine into single data frame
voxdf <- as.data.frame(voxcord)
colnames(voxdf) <- c("x", "y", "z")
voxdf$intensity <- Xp[, , , 459][voxcord]

# set up 3D plot
scene1 <- list(title = "", showticklabels = FALSE, showgrid = TRUE, 
               showline = FALSE, zeroline = FALSE)
marklist <- list(size = 3.5, color = ~intensity, colorscale = 'Viridis', 
                 showscale = TRUE, opacity = 0.5)
fscence <- list(xaxis = scene1, yaxis = scene1, zaxis = scene1)

# generate 3D plot
plot_ly(voxdf, x = ~x, y = ~y, z = ~z, type = "scatter3d", mode = "markers",
        marker = marklist) %>% 
          layout(title = "3D Example of Spleen - Intensity Plot", 
                 scene = fscence)
```

We remove a different number of levels from each object, randomly selecting a 
number of levels between three and six (inclusive) to remove from the first 
mode. The resulting array is ragged: mode A contains a different number of 
levels for each object while modes B and C each contain exactly 28 levels for 
all objects. This removal is meant to simulate motion artifacts from CT scans 
that might make some slices unusable and that might have been removed by a 
quality control algorithm. To continue, we use a Parafac2 model.

```{r, message = FALSE, warning = FALSE}
# set seed
set.seed(500)

# remove levels from mode A for each object and restructure into list of objects
nobj <- length(y)
nremove <- sample(3:6, nobj, replace = TRUE)
Xrag <- vector(mode = "list", length = nobj)
for (i in 1:nobj) {
   ranremove <- c(sample(1:28, nremove[i]))
   image0 <- Xp[-ranremove, , , i]
   Xrag[[i]] <- image0
}
```

## Analysis

We load **cpfa** and initialize the tensor model. First, the data array is a 
ragged four-way array; so we set `model <- "parafac2"`. Second, we initialize 
the number of components to fit by setting `nfac <- c(3, 4)`. Third, we set 
`nstart <- 10`. Fourth, we specify the constraint for each mode using `const`. 
For Parafac2, we set the fourth mode to have non-negative weights.

Next, we initialize classification methods. First, we use 
`method = c("RF", "GBM")` to use RF and GBM. Second, we specify that the problem 
is multiclass, setting `family <- "multinomial"`. Third, we use 10-fold CV in 
our inner training, setting `nfolds <- 10`; and fourth, we set `nrep <- 5` to 
perform five outer train-test splits. Fifth, we set `ratio <- 0.9`. Finally, we 
specify ranges for tuning parameters the number of trees (RF), node size (RF), 
eta (GBM), maximum depth (GBM), subsample (GBM), and number of rounds (GBM), 
wrapping them into the list called `parameters`.

```{r, message = FALSE, warning = FALSE}
# set seed
set.seed(500)

# initialize model 
model <- "parafac2"
nfac <- c(3, 4)
nstart <- 10
const <- c("uncons", "uncons", "uncons", "nonneg")

# initialize classification
method <- c("RF", "GBM")
family <- "multinomial"
nfolds <- 10
nrep <- 5
ratio <- 0.9

# initialize tuning parameters
ntree <- c(400, 600, 800, 1000)
nodesize <- c(4, 8, 16, 32)
eta <- c(0.3, 0.7); max.depth <- c(1, 2)
subsample <- c(0.75, 0.9); nrounds <- c(100, 400)
parameters <- list(ntree = ntree, nodesize = nodesize, eta = eta, 
                   max.depth = max.depth, subsample = subsample, 
                   nrounds = nrounds)

# implement train-test splits with inner k-fold CV to optimize classification
outputR4 <- cpfa(x = Xrag, y = y, model = model, nfac = nfac, nstart = nstart,
                 const = const, method = method, family = family, 
                 nfolds = nfolds, nrep = nrep, ratio = ratio, 
                 parameters = parameters, type.out = "descriptives", 
                 seeds = NULL, plot.out = TRUE, parallel = FALSE, 
                 verbose = FALSE)
```

## Results

We examine classification performance metrics for each model and classifier, 
averaged across train-test splits. We also examine, averaged across train-test 
splits, optimal tuning parameters for each model.

```{r}
# examine classification performance measures - mean across train-test splits
outputR4$descriptive$mean[, 1:2]

# examine optimal tuning parameters - mean across train-test splits
outputR4$mean.opt.tune
```

Accuracy values are between 0.55 and 0.56. The three-component Parafac2 model 
with the RF classifier performed best (i.e., we choose RF because it has fewer
tuning parameters than GBM, in this case). In addition, average optimal tuning 
parameters are available for each classifier. 

Next, we use `plotcpfa`. Looking at plots, we can search for relationships 
between the levels of the modes and model components. In this case, the 
modes A, B, and C correspond to the three dimensions of the organs while the
fourth mode corresponds to the different organ objects.

```{r}
# set seed
set.seed(500)

# plot heatmaps of component weights for optimal model
results4 <- plotcpfa(outputR4, nstart = 10, ctol = 1e-6, verbose = FALSE)
```

We omit interpretations of the three components, given the difficulty 
of visualizing each organ object. However, we do consider box plots of the mode
D weights.

```{r}
# define estimated D weights
mat <- results4$`D Weights`

# set up box plots
groups <- factor(rep(1:length(levels(y)), each = 115), 
                 labels = c("Bladder", "Femur", "Pancreas", "Spleen"))

# generate box plots
for (i in 1:3) {
   boxplot(mat[, i] ~ groups, 
          main = paste("D weights for Component", i), xlab = "Organ",
          ylab = "D Weights", col = "lightblue", border = "darkblue")
}
```

Examining medians, the first component appears to distinguish the spleen from 
other organs; the second component, the femur and spleen from the pancreas and 
bladder; and the third component, the pancreas and spleen from the other organs. 
More work would need to be done to investigate estimated B weights and C weights 
to further interpret these components.

# References

Asisgress, M. (2025). cpfa: Classification with Parallel Factor Analysis. 
R package version 1.2-2, <https://CRAN.R-project.org/package=cpfa>.

Bilic, P., Christ, P., Li, H., Vorontsov, E., Ben-Cohen, A., Kaissis, G., ...
and Menze, B. (2023). The liver tumor segmentation benchmark (lits). Medical 
Image Analysis, 84, 102680.

Fischer, B., Smith, M., and Pau, G. (2025). rhdf5: R Interface to HDF5. 
R package version 2.52.1, <https://bioconductor.org/packages/rhdf5>.

Harshman, R. (1970). Foundations of the PARAFAC procedure: Models and conditions 
for an “explanatory” multi-modal factor analysis. UCLA working papers in 
phonetics, 16(1), 84.

Harshman, R. (1972). PARAFAC2: Mathematical and technical notes. UCLA working 
papers in phonetics, 22(3044), 122215.

Harshman, R. and Lundy, M. (1994). PARAFAC: Parallel factor analysis. 
Computational Statistics and Data Analysis, 18(1), 39-72.

Harshman, R. and Lundy, M. (1996). Uniqueness proof for a family of models 
sharing features of Tucker's three-mode factor analysis and PARAFAC/CANDECOMP. 
Psychometrika, 61(1), 133-154.

Helwig, N. (2025). CMLS: Constrained Multivariate Least Squares.
R package version 1.1, <https://CRAN.R-project.org/package=CMLS>.

Helwig, N. (2025). multiway: Component Models for Multi-Way Data. 
R package version 1.0-7, <https://CRAN.R-project.org/package=multiway>.

Irizarry, R., Gill, A. (2025). dslabs: Data Science Labs.
R package version 0.9.0, <https://CRAN.R-project.org/package=dslabs>.

Kalinowski, T., Allaire, J., and Chollet, F. (2025). keras3: R Interface to 
'Keras'. R package version 1.4.0, <https://CRAN.R-project.org/package=keras3>.

LeCun, Y., Bottou, L., Bengio, Y., and Haffner, P. (2002). Gradient-based 
learning applied to document recognition. Proceedings of the IEEE, 86(11), 
2278–2324.

LeCun, Y., Cortes, C., and Burges, C. (1998). The MNIST database of handwritten 
digits. <http://yann.lecun.com/exdb/mnist/>.

R Core Team (2025). R: A Language and Environment for Statistical Computing. R 
Foundation for Statistical Computing, Vienna, Austria. 
<https://www.R-project.org/>.

Sievert, C. (2020). Interactive web-based data visualization with R, plotly, 
and shiny. Chapman and Hall/CRC.

Xiao, H., Rasul, K., and Vollgraf, R. (2017). Fashion-mnist: a novel image 
dataset for benchmarking machine learning algorithms. arXiv preprint 
arXiv:1708.07747.

Xu, X., Zhou, F., Liu, B., Fu, D., and Bai, X. (2019). Efficient multiple organ 
localization in CT image using 3D region proposal network. IEEE Transactions on 
Medical Imaging, 38(8), 1885-1898.

Yang, J., Shi, R., and Ni, B. (April 2021). Medmnist classification decathlon: 
A lightweight automl benchmark for medical image analysis. In 2021 IEEE 18th 
International Symposium on Biomedical Imaging (ISBI) (pp. 191-195). IEEE.

Yang, J., Shi, R., Wei, D., Liu, Z., Zhao, L., Ke, B., ... and Ni, B. (2023). 
Medmnist v2-a large-scale lightweight benchmark for 2d and 3d biomedical image 
classification. Scientific Data, 10(1), 41.

Yang, X., Xia, D., Kin, T., and Igarashi, T. (2020). Intra: 3d intracranial 
aneurysm dataset for deep learning. In Proceedings of the IEEE/CVF conference 
on computer vision and pattern recognition (pp. 2656-2666).