---
title: "Multiway Classification with Real Data"
author: "Matthew Asisgress"
date: "2025-10-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this demonstration, we apply the R package **cpfa** (Asisgress, 2025) to the 
MNIST dataset (LeCun, Cortes, and Burges, 1998; LeCun et al., 2002). We show how 
to use the package to distinguish between the digits of 2 and 3, a binary
classification problem.

# MNIST

The MNIST dataset consists of 70,000 grayscale images of handwritten digits from 
0 to 9. We use the R package **dslabs** (Irizarry and Gill, 2025) to download 
the dataset. We subset to only 2000 images from the original training set. The
images include 980 images of the digit 2 and 1020 images of the digit 3.

## Download

```{r, message = FALSE, warning = FALSE}
# load library
library(dslabs)

# download MNIST data and subset training set to digits of 2 or 3
mnist <- read_mnist()
inde <- which(mnist$train$labels %in% c(2, 3))
images <- mnist$train$images[inde, ]
labels <- mnist$train$labels[inde]

# restructure data into a three-way array and prepare labels
X0 <- array(images, c(nrow(images), 28, 28))
y0 <- as.factor(as.numeric(as.factor(labels)) - 1)

# subset to first 2000 images
ind <- 1:2e3
X <- X0[ind, , ]
y <- y0[ind]
```

We plot and examine an example of the digit 2 and an example of the digit 3.

```{r}
# plot example of digit 2 and digit 3
par(mfrow = c(1, 2))
pdigit <- function(imat) {
  m <- t(apply(imat, 2, rev))
  image(m, col = gray(seq(0, 1, 0.05)), xaxt = "n", yaxt = "n")
}
pdigit(t(X[which(y == 0)[1], , ])); pdigit(t(X[which(y == 1)[1], , ]))
```

```{r echo = FALSE}
par(mfrow = c(1, 1))
```

## Analysis

We load the R package **cpfa** and initialize the tensor model. First, the data 
array is a regular three-way array, so we set `model <- "parafac"` to use a 
Parafac model. Note that the Parafac2 model can be used for ragged tensors. 
Second, we initialize the number of components to fit for the model by setting 
`nfac <- c(2, 3)` in order to fit both a two-component Parafac model and a 
three-component Parafac model. We set `nstart <- 10` to allow for 10 random 
starts in the Parafac alternating least squares algorithm fit by R package 
**multiway** (Helwig, 2025), upon which R package **cpfa** depends. Third, we 
specify the constraint desired for each array mode using `const`. Note that
numerous constraint options are available; after loading **cpfa**, type 
`const()` in the R console to access a constraint options list provided
through R package **CMLS** (Helwig, 2025). Fourth, we use `cmode <- 1` to 
specify that the classification mode is the first mode of the input array (i.e., 
the mode connected to the class labels).

We next initialize classification methods. First, we use 
`method = c("PLR", "RF")` to employ penalized logistic regression (PLR) and 
random forest (RF) classifiers for this problem. See `help(cpfa)` for other 
options and for references associated with these classifiers. Second, we specify 
that the problem is a binary classification problem by setting 
`family <- "binomial"`. Third, we use 10-fold cross-validation (CV) in our inner 
training, setting `nfolds <- 10`; and fourth, we set `nrep <- 5` to perform five 
outer train-test splits of our data. Fifth, we set `ratio <- 0.9` to specify 
that each outer training set will contain a proportion of 0.9 of the full input 
data while the outer testing set will contain 0.1 of the data. Finally, we 
specify ranges for tuning parameters alpha (PLR), the number of trees (RF), and 
node size (RF), wrapping them into a list called `parameters`.

```{r, message = FALSE, warning = FALSE}
# load library
library(cpfa)

# set seed
set.seed(500)

# initialize Parafac model 
model <- "parafac"
nfac <- c(2, 3)
nstart <- 10
const <- c("uncons", "uncons", "uncons")
cmode <- 1

# initialize classification
method <- c("PLR", "RF")
family <- "binomial"
nfolds <- 10
nrep <- 5
ratio <- 0.9

# initialize tuning parameters
alpha <- seq(0, 1, length = 8)
ntree <- c(400, 600, 800, 1000)
nodesize <- c(4, 8, 16, 32)
parameters <- list(alpha = alpha, ntree = ntree, nodesize = nodesize)

# implement train-test splits with inner k-fold CV to optimize classification
outputR <- cpfa(x = X, y = y, model = model, nfac = nfac, nstart = nstart,
                const = const, cmode = cmode, method = method, 
                family = family, nfolds = nfolds, nrep = nrep, ratio = ratio, 
                parameters = parameters, type.out = "descriptives", 
                seeds = NULL, plot.out = TRUE, parallel = FALSE, 
                verbose = FALSE)

```

## Results

We examine classification performance metrics of error (`err`) and overall
accuracy (`acc`) for each model and for each classifier. We also examine,
averaged across outer train-test splits, the optimal tuning parameters chosen
(i.e., that minimized misclassification error) for each classifier.

```{r}
# examine classification performance measures - median across train-test splits
outputR$descriptive$median[, 1:2]

# examine optimal tuning parameters averaged across train-test splits
outputR$mean.opt.tune
```

Using 2000 images, we can see that classification accuracy values are close to 
0.9 for both two-component and three-component models. The two-component Parafac
model with the RF classifier performed best. In addition, we can see average 
optimal tuning parameters for each classifier. Note that PLR optimized lambda 
internally.

Next, we use the function `plotcpfa`. The function does two things: (1) it fits
the optimal model among those fit using function `cpfa`; and (2) it plots the 
component weights from the non-classification modes. Note that, for the Parafac2
model, the function does not plot A mode weights as these are allowed to vary
within each level of the classification mode. From the component weights, we can
search for relationships between the levels of mode A or B and the model 
components. In this case, mode A corresponds to the horizontal axis of the 
images (when viewing the images as those above) while mode B corresponds to the
vertical axis of the images.

```{r}
# set seed
set.seed(500)

# plot heatmaps of component weights for optimal model
results <- plotcpfa(outputR, nstart = 10, ctol = 1e-6, verbose = FALSE)
```

Examining the first component for the A mode (horizontal), weights are stronger
between levels 5-24. Looking at the B mode (vertical), a similar pattern is 
seen: weights are stronger in the middle between levels 4-25. Then, whether 
viewed from the A mode or B mode perspective, this first component appears to be 
a general component showing the presence of non-zero values, distinguishing 
empty outer parts of images from inner parts. Likely, the 2 and 3 digits have
systematic differences in the presence of non-zero values in these areas.

Examining the second component for the A mode, weights are stronger near and
around level 14 where there appears to be a greater presence of zero values in 
digit 3 compared to digit 2 near this level. A similar pattern is seen in the
B mode around levels 18 and 19 with more non-zero values for 2, compared to 3,
around these levels. Taken together between the modes, this component could be
identifying the closed circle present in the digit 2, which differs from the 
open or half circle seen in digit 3.

# References

Asisgress, M. (2025). cpfa: Classification with Parallel Factor Analysis. 
R package version 1.2-2, <https://CRAN.R-project.org/package=cpfa>.

Helwig, N. (2025). CMLS: Constrained Multivariate Least Squares.
R package version 1.1, <https://CRAN.R-project.org/package=CMLS>.

Helwig, N. (2025). multiway: Component Models for Multi-Way Data. 
R package version 1.0-7, <https://CRAN.R-project.org/package=multiway>.

Irizarry, R., Gill, A. (2025). dslabs: Data Science Labs.
R package version 0.9.0, <https://CRAN.R-project.org/package=dslabs>.

LeCun, Y., Bottou, L., Bengio, Y., and Haffner, P. (2002). Gradient-based 
learning applied to document recognition. Proceedings of the IEEE, 86(11), 
2278–2324.

LeCun, Y., Cortes, C., and Burges, C. (1998). The MNIST database of handwritten 
digits. <http://yann.lecun.com/exdb/mnist/>.

R Core Team (2025). R: A Language and Environment for Statistical Computing. R 
Foundation for Statistical Computing, Vienna, Austria. 
<https://www.R-project.org/>.